
---
title: 'STAT2003: Assignment 1'
author: "Samantha Clements and Kafula Chewe"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<<<<<<< HEAD
```{r Install Packages, include=FALSE}
# install.packages("ISLR", dependencies = TRUE)
# install.packages("caret", dependencies = TRUE)
# install.packages("e1071", dependencies = TRUE)
# install.packages("randomForest", dependencies = TRUE)
# install.packages("DiagrammeR", dependencies = TRUE)

```
=======

>>>>>>> b556a26 (DOE diagram)


```{r Library, include=FALSE}
library(ISLR)
library(e1071)
library(randomForest)
library(DiagrammeR)
library(caret)

```

# Machine Learning Algorithms Comparison

This experiment evaluates the effectiveness of four machine learning methods: Logistic Regression (LR), Random Forest (RF), Naive Bayes (NB), and Support Vector Machine (SVM). The comparison involved the models ability to accurately and precisely predict credit card defaults using the Default data set from the ISLR package in R (James et al. 2013). 

(justification of response)

The experiment is subject to the four principles of experimental design (Montgomery 2020). Randomisation was conducted through the random assignment of observations to cross-validation (CV) fold, with set.seed(123), to ensure reproducibility across all models.



```{r DOE Diagram, echo = FALSE }
# sources:
# https://www.sesync.org/resources/creating-visualizations-diagrammer 
# clusters: https://graphviz.org/pdf/dotguide.pdf

DOE_Diagram <- grViz("digraph{
         
                     graph[layout = dot, rankdir = LR, splines = line] # Left to Right
                     
                     node[shape = rectangle, style = filled]
                     
                    # Central box
                    
                    centre_box [label = '\n\n\nDOE\n\n\n\n', 
                    shape = box, fillcolor = lightblue, width = 3]\
                    
                    # Inputs
                    
                    subgraph cluster_input {
                    
                    
                    X1 [label = 'X1: Model Type: \nLogistic Regression,\nRandom Forest,\nNaive Bayes,\nSupport Vector Machine']
                    
                    {rank = same
                    edge [constraint = false, style = invis]
                    
                    # Vertically stacked
                    X1 }
                    
                    }
                    
                    
                    # Noise
                    subgraph cluster_noise {
                    
                    N1 [label = 'N1: Random Sampling Variation']
                    N2 [label = 'N2: Model Randomness']
                    N3 [label = 'N3: CV Fold Assignment Randomness']
                    
                    {edge [style = invis]
                    
                    # Horizontally stacked
                    N1 -> N2 -> N3}
                    
                    }
                    
                    
                    # Constants
                    subgraph cluster_constants {
                    
                    
                    
                    C1 [label = 'C1: Same random seed - set.seed(000)']
                    C2 [label = 'C2: Same Dataset - Default (ISLR)']
                    C3 [label = 'C3: CV - 10 folds, 3 repeats']
                    C4 [label = 'C4: Model Hyperparameters']
                    
                    
                    {edge [style = invis]
                    
                    # Horizontally stacked
                    C1 -> C2 -> C3 -> C4}
                    }
                    
                  
                    # Outputs
                    subgraph cluster_outputs {
                    
                    Y1 [label = 'Y1: AUC-ROC']
                    Y2 [label = 'Y2: Precision']
                    Y3 [label = 'Y3: Recall']
                    
                    {rank = same
                    edge [constraint = false, style = invis]
                    
                    # Vertically stacked
                    Y1 -> Y2 -> Y3}
                    
                    }
                   
                  
                    # Arrows
                    
                    edge [arrowhead = normal, style = solid]
                    X1 -> centre_box [minlen = 2]
                   
                    edge [arrowhead = normal, style = solid]
                    centre_box -> Y1 [minlen = 2.5]
                    centre_box -> Y2
                    centre_box -> Y3
                    
                    edge [constraint = false, arrowhead = normal, style = solid]
                    
                    N1 -> centre_box:s [minlen = 3] 
                    N2 -> centre_box:s [minlen = 3] 
                    N3 -> centre_box:s [minlen = 3]
                    
                    edge [constraint = false, arrowhead = normal, style = solid]
                    C1 -> centre_box:n [minlen = 3]
                    C2 -> centre_box:n [minlen = 3]
                    C3 -> centre_box:n [minlen = 3]
                    C4 -> centre_box:n [minlen = 3]
                    
                    
                     }")
 
DOE_Diagram
```

QUESTION 2
To conduct this experiment a Randomised Block Design (RBD) was used to compare the four different machine learning algorithms. 
The experimental factor was the algorithm with the four levels being represented by the the machine learning algorithms: Logistic Regression, Random Forrest, Naive Bayes and Support Vector Machine. The response variable is ROC AUC which explains how good the algorithms are determining whether someone defaulted or not. To control background variation due to random partitioning, that is, one split making one algorithm look better than another the experiment contains repeated 10-fold cross validation with 3 repeats, therefore, creating 30 resampling blocks. To ensure that it was a fair comparison, within each block all four algorithms were tested under the same split. 

The results below show differences in the ROC averages across all four algorithms. Logistic Regression had the highest ROC with Naive Bayes being closely behind while Random forrest was lower on average and SVM had the most variability. The boxplot further shows that Logistic Regression is not only stronger on average but is stable across all blocks. 

On the basis of ROC mean, the experiment suggests that logistic regression provides the strongest overall best performance.

```{r Data Obtain and Inspect, echo = FALSE}
set.seed(000)
#Obtaining and inspecting the data
data(Default)
Default$default <- factor(Default$default, levels = c("No", "Yes"))
head(Default)

table(Default$default)

```

```{r Experiment Control, include = FALSE}
install.packages("DiagrammeR", dependencies = TRUE)
```

```{r DOE Diagram}
# sources:
# https://www.sesync.org/resources/creating-visualizations-diagrammer 
# clusters: https://graphviz.org/pdf/dotguide.pdf

# install.packages("DiagrammeR", dependencies = TRUE)
library(DiagrammeR)

DOE_Diagram <- grViz("digraph{
         
                     graph[layout = dot, rankdir = LR, splines = line] # Left to Right
                     
                     node[shape = rectangle, style = filled]
                     
                    # Central box
                    
                    centre_box [label = '\n\n\nDOE\n\n\n\n', 
                    shape = box, fillcolor = lightblue, width = 3]\
                    
                    # Inputs
                    
                    subgraph cluster_input {
                    
                    
                    X1 [label = 'X1: Model Type: \nLogistic Regression,\nRandom Forest,\nNaive Bayes,\nSupport Vector Machine']
                    
                    {rank = same
                    edge [constraint = false, style = invis]
                    
                    # Vertically stacked
                    X1 }
                    
                    }
                    
                    
                    # Noise
                    subgraph cluster_noise {
                    
                    N1 [label = 'N1: Random Sampling Variation']
                    N2 [label = 'N2: Model Randomness']
                    N3 [label = 'N3: CV Fold Assignment Randomness']
                    
                    {edge [style = invis]
                    
                    # Horizontally stacked
                    N1 -> N2 -> N3}
                    
                    }
                    
                    
                    # Constants
                    subgraph cluster_constants {
                    
                    
                    
                    C1 [label = 'C1: Same random seed - set.seed(000)']
                    C2 [label = 'C2: Same Dataset - Default (ISLR)']
                    C3 [label = 'C3: CV - 10 folds, 3 repeats']
                    C4 [label = 'C4: Model Hyperparameters']
                    
                    
                    {edge [style = invis]
                    
                    # Horizontally stacked
                    C1 -> C2 -> C3 -> C4}
                    }
                    
                  
                    # Outputs
                    subgraph cluster_outputs {
                    
                    Y1 [label = 'Y1: AUC-ROC']
                    Y2 [label = 'Y2: Precision']
                    Y3 [label = 'Y3: Recall']
                    
                    {rank = same
                    edge [constraint = false, style = invis]
                    
                    # Vertically stacked
                    Y1 -> Y2 -> Y3}
                    
                    }
                   
                  
                    # Arrows
                    
                    edge [arrowhead = normal, style = solid]
                    X1 -> centre_box [minlen = 2]
                   
                    edge [arrowhead = normal, style = solid]
                    centre_box -> Y1 [minlen = 2.5]
                    centre_box -> Y2
                    centre_box -> Y3
                    
                    edge [constraint = false, arrowhead = normal, style = solid]
                    
                    N1 -> centre_box:s [minlen = 3] 
                    N2 -> centre_box:s [minlen = 3] 
                    N3 -> centre_box:s [minlen = 3]
                    
                    edge [constraint = false, arrowhead = normal, style = solid]
                    C1 -> centre_box:n [minlen = 3]
                    C2 -> centre_box:n [minlen = 3]
                    C3 -> centre_box:n [minlen = 3]
                    C4 -> centre_box:n [minlen = 3]
                    
                    
                     }")
 
DOE_Diagram
```


```{r}
(DOE diagram)
ctrl <- trainControl(method = "repeatedcv",
                     number = 10,
                     repeats = 3,          
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = TRUE)
```



```{r Model Training, echo = FALSE}
#Training the models
model_lr  <- train(default ~ ., data = Default, method = "glm",
                   family = "binomial", trControl = ctrl, metric = "ROC")

model_rf  <- train(default ~ ., data = Default, method = "rf",  
                   trControl = ctrl, metric = "ROC",
                   tuneLength = 3)  

model_nb  <- train(default ~ ., data = Default, method = "nb",
                   trControl = ctrl, metric = "ROC")

model_svm <- train(default ~ ., data = Default, method = "svmLinear",
                   trControl = ctrl, preProcess = c("center", "scale"),
  tuneLength = 3)
```

```{r Models Listed, include = FALSE}
#Creating a list of all four models
models <- list(LogisticRegression = model_lr,
               RandomForest = model_rf,
               NaiveBayes = model_nb,
               SVM = model_svm)
```

```{r Resampling Summary, echo = FALSE}
#Summary of resampling results
res <- resamples(models)

summary(res)
```


```{r ROC Plot, echo = FALSE}
#Creates resamples that compares ROC across 30 resamples 
bwplot(res, metric = "ROC",
       main = "ROC AUC across 30 resamples (3×10 repeated CV)")
```


QUESTION 3
Logistic Regression Suggestion (Limitation based on one dataset)

```{r References, include = FALSE}

# Chicago 17th Author-Date. NOTE: I will update .bib file later
#

## (Montgomery 2020)
# Montgomery, Douglas C. 2020. Design and Analysis of Experiments. 1st ed. Hoboken, NJ: Wiley.

## (Kleijnen 2008)
# Kleijnen, Jack P. C. 2008. Design and Analysis of Simulation Experiments. 1st ed. New York: Springer-Verlag.
#



## (James et al. 2013)
# James, Gareth, Daniela Witten, Trevor Hastie, and Robert Tibshirani. 2013. An Introduction to Statistical Learning. Vol. 112. New York: Springer.



```

